---
title: BD2HAN Rat RNAseq Deconvolution
output:
  html_document: default
---

### **Overview**

Data was processed using Nextflow 'brucemoran/RNAseq_kallisto' and 'brucemoran/aracne-ap_viper' pipelines. `Sleuth` objects were parsed using scripts available in this project to generate RData files used in this document.

```{r echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
##working dir is location of Rmd, here BM/deconv
load("../data/BD2HAN_2_RNAseqR/RData/BD2HAN_2.DE_ready.RData")
load("../data/ARACNe/BD2HAN_2.parse_inputs.RData")
metadata <- readr::read_csv("../data/metadata/BDII_Han_Rat2_metadata.TISSUE_ARM_immune_090321.csv")

```

### **Data formatting**

We require HGNC (human) annotations for use in the [`immunedeconv` R package](https://github.com/icbi-lab/immunedeconv). This allows us to run 4 deconvolution methods: `quantiseq`, `mcp_counter`, `xcell` and `epic`.

```{r echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
library(immunedeconv)
library(tidyverse)
library(ComplexHeatmap)

outdir <- "output/deconv"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

if(!file.exists(paste0(outdir,"/deconv.RData"))){
  tpm_tb_h <- dplyr::select(.data = tpm_tb, -external_gene_name) %>%
              dplyr::left_join(tx2gene, .) %>%
              dplyr::select(-target_id , -ensembl_gene_id, -external_gene_name) %>%
              dplyr::distinct() %>%
              na.omit() %>%
              dplyr::filter(!hsapiens_homolog_associated_gene_name %in% "")

  ##unique names
  tpm_tb_hn <- names(which(table(tpm_tb_h$hsapiens_homolog_associated_gene_name)==1))

  tpm_tb_h_gat <- RNAseqon::group_agg_two(f = tpm_tb_h, pattern="R") %>%
                  dplyr::filter(hsapiens_homolog_associated_gene_name %in% !!tpm_tb_hn) %>%
                  dplyr::select(-hsapiens_homolog_ensembl_gene) %>%
                  as.data.frame() %>%
                  tibble::column_to_rownames("hsapiens_homolog_associated_gene_name")

  deconv_list <- list(1:4)
  deconv_list[[1]] <- immunedeconv::deconvolute(tpm_tb_h_gat, "quantiseq", tumor = TRUE)
  deconv_list[[2]] <- immunedeconv::deconvolute(tpm_tb_h_gat, "mcp_counter", tumor = TRUE)
  deconv_list[[3]] <- immunedeconv::deconvolute(tpm_tb_h_gat, "xcell", tumor = TRUE)
  deconv_list[[4]] <- immunedeconv::deconvolute(tpm_tb_h_gat, "epic", tumor = TRUE)
  names(deconv_list) <- c("quantiseq", "mcp_counter", "xcell", "epic")

  ##plot list
  plot_list <- lapply(seq_along(deconv_list), function(f){

    decon_char <- deconv_list[[f]] %>%
                  dplyr::filter(!cell_type %in% "uncharacterized cell")

    ggp <- decon_char %>%
      dplyr::filter(!cell_type %in% "uncharacterized cell") %>%
      gather(sample, fraction, -cell_type) %>%
      # plot as stacked bar chart
      ggplot(aes(x = sample, y = fraction, fill = cell_type)) +
        geom_bar(stat = 'identity') +
        coord_flip() +
        scale_x_discrete(limits = rev(levels(decon_char))) +
        ggtitle(paste0("Deconvolution with: ", names(deconv_list)[f]),
                subtitle = "uncharacterised cells removed")


    res_m <- as.matrix(decon_char[,-1])
    rownames(res_m) <- unlist(decon_char[,1])

    ##metadata to correct format and cols for annotation
    metadat_df <- metadata %>% dplyr::filter(sample %in% colnames(res_m)) %>%
                            dplyr::arrange(match(sample, colnames(res_m))) %>%
                            dplyr::select(sample, IMMUNE, TUMOUR_TYPE, TISSUE, CARE_ARM) %>%
                            as.data.frame(stringsAsFactors = TRUE) %>%
                            tibble::column_to_rownames("sample") %>%
                            dplyr::mutate(across(everything(), factor))

    ##annotation for heatmap
    hann <- HeatmapAnnotation(df = metadat_df, na_col = "white",
                              col = list(IMMUNE = c("EXCL" = "orange", "HIGH" = "darkred",
                                                  "LOW" = "dodgerblue", "MED" = "grey",
                                                  "RICH" = "red", "POOR" = "lightblue"),
                                         TUMOUR_TYPE = c("ENDO" = "forestgreen", "NOT" = "white",
                                                  "SER" = "yellow", "SQC" = "pink"),
                                         TISSUE = c("N" = "green", "T" = "black"),
                                         CARE_ARM = c("LG" = "orange", "STD" = "blue")))

    hmp <- Heatmap(res_m,
                   name = names(deconv_list)[f],
                   column_names_gp = gpar(fontsize = 10),
                   bottom_annotation = hann,
                   column_title =  paste0("Deconvolution using: ", names(deconv_list)[f]))

    return(list(barplot = ggp, heatmap = hmp, heatmap_data = res_m))
  })
  names(plot_list) <- names(deconv_list)
  save(deconv_list, plot_list, tpm_tb_h, tpm_tb_hn, tpm_tb_h_gat,
       file = paste0(outdir,"/deconv.RData"))
} else {
  load(paste0(outdir,"/deconv.RData"))
}
```
******
\newpage

## **BD2HAN Validation Data : Plots**

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 10, comment = NA}
lapply(plot_list, function(f){
  f[2][[1]]
})
```

BD2Han_1 Data generated by Nextflow 'RNAseq_kallisto' into 'Sleuth' object (so)

```{r echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
##working dir is location of Rmd, here BM/deconv
load("../data/BD2HAN_1_data/sleuthObject.full.RData")
ls()
so_t <- RNAseqon::so_obs_raw_out(so)
tpm_tb_1 <- so_t$obs_raw_tpm$wide[,-3]
tpm_tb_1_h <- dplyr::select(.data = tpm_tb_1, -external_gene_name) %>%
              dplyr::left_join(tx2gene, .) %>%
              dplyr::select(-target_id , -ensembl_gene_id, -external_gene_name) %>%
              dplyr::distinct() %>%
              na.omit() %>%
              dplyr::filter(!hsapiens_homolog_associated_gene_name %in% "")

##unique names
nn <- names(which(table(tpm_tb_1_h$hsapiens_homolog_associated_gene_name)>1))

tpm_tb_1_h_gat <- RNAseqon::group_agg_two(f = tpm_tb_1_h, pattern = "hsapiens") %>%
                  dplyr::filter(!hsapiens_homolog_associated_gene_name %in% nn) %>%
                  dplyr::select(-hsapiens_homolog_ensembl_gene) %>%
                  dplyr::distinct() %>%
                  as.data.frame() %>%
                  tibble::column_to_rownames("hsapiens_homolog_associated_gene_name")

deconv_list_1 <- list(1:4)
deconv_list_1[[1]] <- immunedeconv::deconvolute(tpm_tb_1_h_gat, "quantiseq", tumor = TRUE)
deconv_list_1[[2]] <- immunedeconv::deconvolute(tpm_tb_1_h_gat, "mcp_counter", tumor = TRUE)
deconv_list_1[[3]] <- immunedeconv::deconvolute(tpm_tb_1_h_gat, "xcell", tumor = TRUE)
deconv_list_1[[4]] <- immunedeconv::deconvolute(tpm_tb_1_h_gat, "epic", tumor = TRUE)
names(deconv_list_1) <- c("quantiseq", "mcp_counter", "xcell", "epic")

##plot list
plot_list_1 <- lapply(seq_along(deconv_list_1), function(f){

  decon_char <- deconv_list_1[[f]] %>%
                dplyr::filter(!cell_type %in% "uncharacterized cell")

  ggp <- decon_char %>%
    dplyr::filter(!cell_type %in% "uncharacterized cell") %>%
    gather(sample, fraction, -cell_type) %>%
    # plot as stacked bar chart
    ggplot(aes(x = sample, y = fraction, fill = cell_type)) +
      geom_bar(stat = 'identity') +
      coord_flip() +
      scale_x_discrete(limits = rev(levels(decon_char))) +
      ggtitle(paste0("Deconvolution with: ", names(deconv_list_1)[f]),
              subtitle = "uncharacterised cells removed")


  res_m <- as.matrix(decon_char[,-1])
  rownames(res_m) <- unlist(decon_char[,1])

  ##metadata to correct format and cols for annotation
  metadat_df <- metadata %>% dplyr::filter(sample %in% colnames(res_m)) %>%
                          dplyr::filter(Tissue %in% "Normal",
                                        Intervention %in% c("RES", "LIR")) %>%
                          dplyr::arrange(match(sample, colnames(res_m))) %>%
                          dplyr::select(sample, Tissue, Diet, Intervention) %>%
                          as.data.frame(stringsAsFactors = TRUE) %>%
                          tibble::column_to_rownames("sample") %>%
                          dplyr::mutate(across(everything(), factor))

  ##annotation for heatmap
  hann <- HeatmapAnnotation(df = metadat_df, na_col = "white",
                            col = list(Intervention = c("LIR" = "orange", "RES" = "blue")))
  res_m <- res_m[,colnames(res_m) %in% rownames(metadat_df)]
  hmp <- Heatmap(res_m,
                 name = names(deconv_list_1)[f],
                 column_names_gp = gpar(fontsize = 10),
                 bottom_annotation = hann,
                 column_title =  paste0("Deconvolution using: ", names(deconv_list_1)[f]))

  return(list(barplot = ggp, heatmap = hmp, heatmap_data = res_m))
})
names(plot_list_1) <- names(deconv_list_1)
save(deconv_list_1, plot_list_1, tpm_tb_1_h, tpm_tb_1_h_gat,
     file = paste0(outdir,"/deconv_1.RData"))
```

******

\newpage

## **BD2HAN Initital Dataset : Plots**

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 10, comment = NA}
lapply(plot_list_1, function(f){
  f[2][[1]]
})
```
