---
title: BD2HAN Rat RNAseq Normal Sample Deconvolution
output:
  html_document: default
---

### **Overview**

Data was processed using Nextflow 'brucemoran/RNAseq_kallisto' and 'brucemoran/aracne-ap_viper' pipelines. `Sleuth` objects were parsed using scripts available in this project to generate RData files used in this document.

```{r echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
##working dir is location of Rmd, here BM/deconv
load("../data/BD2HAN_2_RNAseqR/RData/BD2HAN_2.DE_ready.RData")
load("../data/ARACNe/BD2HAN_2.parse_inputs.RData")
metadata <- readr::read_csv("../data/metadata/BDII_Han_Rat2_metadata.TISSUE_ARM_immune_090321.csv")
```

### **Data formatting**

We require HGNC (human) annotations for use in the [`immunedeconv` R package](https://github.com/icbi-lab/immunedeconv). This allows us to run 4 deconvolution methods: `quantiseq`, `mcp_counter`, `xcell` and `epic`.

```{r echo = FALSE, include = FALSE, message = FALSE, warning = FALSE}
library(immunedeconv)
library(tidyverse)
library(ComplexHeatmap)

outdir <- "output/deconv_normal"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

if(!file.exists(paste0(outdir,"/deconv_normals.RData"))){

  dnorm <- metadata[grep("N_", metadata$TISSUE_ARM), ]
  tpm_norm_tb <- tpm_tb[,colnames(tpm_tb) %in% c("external_gene_name",
                                             "ensembl_gene_id",
                                             dnorm$sample)]
  tpm_norm_tb_h <- dplyr::select(.data = tpm_norm_tb, -external_gene_name) %>%
              dplyr::left_join(tx2gene, .) %>%
              dplyr::select(-target_id , -ensembl_gene_id, -external_gene_name) %>%
              dplyr::distinct() %>%
              na.omit() %>%
              dplyr::filter(!hsapiens_homolog_associated_gene_name %in% "")

  ##unique names
  tpm_norm_tb_hn <- names(which(table(tpm_norm_tb_h$hsapiens_homolog_associated_gene_name)==1))

  tpm_norm_tb_h_gat <- RNAseqon::group_agg_two(f = tpm_norm_tb_h, pattern="R") %>%
                  dplyr::filter(hsapiens_homolog_associated_gene_name %in% !!tpm_norm_tb_hn) %>%
                  dplyr::select(-hsapiens_homolog_ensembl_gene) %>%
                  as.data.frame() %>%
                  tibble::column_to_rownames("hsapiens_homolog_associated_gene_name")

  deconv_norm_list <- list(1:4)
  deconv_norm_list[[1]] <- immunedeconv::deconvolute(tpm_norm_tb_h_gat, "quantiseq", tumor = TRUE)
  deconv_norm_list[[2]] <- immunedeconv::deconvolute(tpm_norm_tb_h_gat, "mcp_counter", tumor = TRUE)
  deconv_norm_list[[3]] <- immunedeconv::deconvolute(tpm_norm_tb_h_gat, "xcell", tumor = TRUE)
  deconv_norm_list[[4]] <- immunedeconv::deconvolute(tpm_norm_tb_h_gat, "epic", tumor = TRUE)
  names(deconv_norm_list) <- c("quantiseq", "mcp_counter", "xcell", "epic")

  ##plot list
  plot_norm_list <- lapply(seq_along(deconv_norm_list), function(f){

    decon_char <- deconv_norm_list[[f]] %>%
                  dplyr::filter(!cell_type %in% "uncharacterized cell")

    ggp <- decon_char %>%
      dplyr::filter(!cell_type %in% "uncharacterized cell") %>%
      gather(sample, fraction, -cell_type) %>%
      # plot as stacked bar chart
      ggplot(aes(x = sample, y = fraction, fill = cell_type)) +
        geom_bar(stat = 'identity') +
        coord_flip() +
        scale_x_discrete(limits = rev(levels(decon_char))) +
        ggtitle(paste0("Deconvolution with: ", names(deconv_norm_list)[f]),
                subtitle = "uncharacterised cells removed")


    res_m <- as.matrix(decon_char[,-1])
    rownames(res_m) <- unlist(decon_char[,1])

    ##metadata to correct format and cols for annotation
    metadat_df <- metadata %>% dplyr::filter(sample %in% colnames(res_m)) %>%
                            dplyr::arrange(match(sample, colnames(res_m))) %>%
                            dplyr::select(sample, IMMUNE, TUMOUR_TYPE, TISSUE, CARE_ARM) %>%
                            as.data.frame(stringsAsFactors = TRUE) %>%
                            tibble::column_to_rownames("sample") %>%
                            dplyr::mutate(across(everything(), factor))

    ##annotation for heatmap
    hann <- HeatmapAnnotation(df = metadat_df, na_col = "white",
                              col = list(IMMUNE = c("EXCL" = "orange", "HIGH" = "darkred",
                                                  "LOW" = "dodgerblue", "MED" = "grey",
                                                  "RICH" = "red", "POOR" = "lightblue"),
                                         TUMOUR_TYPE = c("ENDO" = "forestgreen", "NOT" = "white",
                                                  "SER" = "yellow", "SQC" = "pink"),
                                         TISSUE = c("N" = "green", "T" = "black"),
                                         CARE_ARM = c("LG" = "orange", "STD" = "blue")))

    hmp <- Heatmap(res_m,
                   name = names(deconv_norm_list)[f],
                   column_names_gp = gpar(fontsize = 10),
                   bottom_annotation = hann,
                   column_title =  paste0("Deconvolution using: ", names(deconv_norm_list)[f]))

    return(list(barplot = ggp, heatmap = hmp, heatmap_data = res_m))
  })
  names(plot_norm_list) <- names(deconv_norm_list)
  save(deconv_norm_list, plot_norm_list, tpm_norm_tb_h, tpm_norm_tb_hn, tpm_norm_tb_h_gat,
       file = paste0(outdir,"/deconv_normals.RData"))
} else {
  load(paste0(outdir,"/deconv_normals.RData"))
}
```
******
\newpage

## **BD2HAN Validation Data : Plots**

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 10, comment = NA}
lapply(plot_norm_list, function(f){
  f[2][[1]]
})
```
